---
title: "Synergy screen example"
author: "Leiv Rønneberg"
date: "05/09/2022"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Synergy screen example}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The `synergyscreen` provides a work flow for data from big drug
combination screens, where multiple drugs are tested in combination on
multiple cell lines. It takes as input a list of experiments, each entry
being a list containing the necessary elements needed for a call to the
main regression function `bayesynergy`.

# Melanoma cell line example

Included in the package is the result of a `synergyscreen` run of 583
drug combinations on the A-375 human melanoma cell line from
@ONeil_2016. The `synergyscreen` object is a list with two entries, a
dataframe with parameter estimates from each experiment, and a list
entitled `failed` -- containing experiments that either failed
completely to process, or had an unsatisfactory fit.

```{r synergyscreen_load, cache = T}
library(bayesynergy)
data("ONeil_A375")
length(ONeil_A375$failed)
```

We see that the dataset has two experiments that failed to process,
during an initial run of `synergyscreen`. There's a multitude of reasons
why an experiment might fail to process, it could be an input error,
initialization problems or problems with the parallel processing.

The entries of `failed` are themselves lists, each containing the
necessary information to process through the `bayesynergy` function

```{r synergyscreen_load2, cache = T}
failed_experiment = ONeil_A375$failed[[1]]
names(failed_experiment)
```

```{r, echo = F,cache=T}
colnames(failed_experiment$x) = failed_experiment$drug_names
colnames(failed_experiment$y) = "viability"
```

```{r,cache=T}
head(cbind(failed_experiment$y,failed_experiment$x))
```

We can rerun experiments that failed to process, by simply passing the
returned `synergyscreen` object back into the function. Note that we
turn of the default options of saving each fit and plotting everything,
and set `method = "vb"` indicating we use variational inference to fit
the model.

```{r synergyscreen_fit, cache = T, warning = F}
fit_screen = synergyscreen(ONeil_A375, save_raw = F, save_plots = F, parallel = F, 
                           bayesynergy_params = list(method = "vb"))
```

# Visualization

We can also plot the result of the screen:

```{r synergyscreen, cache = T, warning = FALSE, message = FALSE, fig.dim = c(8,8), fig.show="hold", results="hide", fig.keep ="all"}
plot(fit_screen)
```

# References
