---
title: "Advanced topics"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Advanced topics}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Diagnosing errors and warnings

Sometimes, the `bayesynergy` function may return with a warning.
Ideally, we don't want any warnings at all, and they should be examined
closely, as posterior samples could be unreliable. Usually, the warning
will tell the user how to fix the problem at hand, e.g. by running the
chains for longer (set `iter` higher), or setting `adapt_delta` higher.
See [<https://mc-stan.org/misc/warnings.html>] for some general tips.

## Divergent Transitions

Most commonly, the sampler might complain about divergent transitions.
The warning will typically look like this:

```         
## Warning: There were 2316 divergent transitions after warmup. See
## http://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
## to find out why this is a problem and how to eliminate them.
```

This is indicative of a posterior geometry that is tricky to explore. In
the case where there is only a few divergent transitions, the usual
trick is to set `adapt_delta` to a higher value, i.e. larger than 0.9
which is the default. This can be done through the `control` option in
the `bayesynergy` call:

```{r, eval = F}
fit = bayesynergy(y, x, control = list(adapt_delta = 0.99))
```

However, the case above, where there are 2316 divergent transitions, is
indicative of a misspecified model. In my experience, this can happen
for a few reasons.

-   Estimating 'flat' monotherapies
    -   i.e. when the parameter $l$ is close to one. This can have the
        effect of making the other monotherapy parameters
        unidentifiable.
    -   this can usually be alleviated by setting
        `lower_asymptotes = FALSE` in the call. Unless one is
        specifically interested in these parameters, there are no reason
        to estimate them -- the model fit will typically still be good
        without them.
-   Estimating $l$ in the heteroscedastic model
    -   the model can struggle in this setting, particularly if there
        are none or few replicates.
    -   choose a homoscedastic model instead.
-   'Wrong' $\lambda$ value
    -   sometimes setting $\lambda$ much lower than the initial setting
        can help with a better fit. This is particularly true if
        viability scores close to zero (or negative) are truncated or
        set to exactly zero.

# Robustness against outliers

We provide a version of the model that is more robust against inevitable
outliers in dose=response data. This is done by swapping out two
components of the model formulation, the likelihood and the prior for
kernel hyperparameters.

For the likelihood, we utilise the log-Pareto-tailed Normal (LPTN)
distribution with mean $\mu$ and standard deviation $\sigma$, as
described in @Gagnon2020 which takes the form $$
f(x)=
\begin{cases}
\frac{1}{\sigma}\phi\left(\frac{x-\mu}{\sigma}\right) & \text{if } \vert\frac{x-\mu}{\sigma}\vert \leq \tau \\
\frac{1}{\sigma}\phi(\tau)\frac{\tau}{\vert\frac{x-\mu}{\sigma}\vert}\left(\frac{\log (\tau)}{\log(\vert\frac{x-\mu}{\sigma}\vert)}\right)^{\lambda+1} & \text{if } \vert\frac{x-\mu}{\sigma}\vert > \tau
\end{cases}
$$ where $\phi$ denotes the standard normal pdf. The parameters
$(\tau,\lambda)$ are controlled by the user-specified hyperparameter
$\rho \in (2\Phi(1)-1,1)$ as $$
\tau=\Phi^{-1}((1+\rho)/2),  \ \ \ \ \lambda=2(1-\rho)^{-1}\phi(\tau)\tau\log(\tau)
$$

The robust likelihood can be utilized by setting `robust = T` when
calling the`bayesynergy` function. The user-specified parameter $\rho$
can be set by the `rho` argument, by default set to $0.9$.

For the kernel, we recommend utilising the Matérn kernel with a
Penalized Complexity (PC) prior on the kernel hyperparameters. The PC
prior for the Matern covariance was described in @Fuglstad2018, and
strongly encourages small deviations from the null model, i.e. high
probability of an interaction term being zero. The PC prior for the
kernel hyperparameters $(\sigma_f^2,\ell)$ (in the 2-dimensional case)
takes the form $$
\pi(\sigma_f^2,\ell)=\tilde{\lambda}_1\tilde{\lambda}_2\ell^{-2}\sigma_f^{-1}\exp\left(-\tilde{\lambda}_1\ell^{-1}-\tilde{\lambda}_2\sigma_f\right),
$$ where $(\tilde{\lambda}_1,\tilde{\lambda}_2)$ are set to reflect
prior beliefs $P(\ell < \ell_0)=\alpha_1$ and
$P(\sigma_f^2 > \sigma^2_{f0})=\alpha_2$ by $$
\tilde{\lambda}_1=-\log(-\alpha_1)\ell \ \ \ \ \tilde{\lambda}_2=-\frac{\log(\alpha_2)}{\sigma_{f0}^2},
$$ where $(\ell_0,\alpha_1,\sigma_{f0}^2,\alpha_2)$ are hyperparameters
that are set by the user. The PC prior can be enabled by setting
`pcprior = T` when calling the `bayesynergy` function, and
hyperparameters specified by the argument `pcprior_hypers`. By default,
we recommend $(\ell_0,\alpha_1,\sigma_{f0}^2,\alpha_2)=(1,0.1,1,0.2)$.

# Including controls

The positive and negative controls essentially control the
signal-to-noise ratio in cell viability assays. If the user has access
to these, they can be included in the model to help calibrate the
posterior distribution -- particularly in the case with zero replicates.

Let $\xi^-_k$ and $\xi^+_l$ denote the negative and positive controls
for $k=1,\ldots,n_-$ and $l=1,\ldots,n_+$. These measurements are raw
readings from the plate and are used to calculate cell viability. For an
additional well, treated with drug concentration $\mathbf{x}_i$, we
denote the raw output by $\xi_i$, and calculate cell viability for this
well by the formula: $$
y_i = \frac{\xi_i-\tilde{\xi^+}}{\tilde{\xi^-}-\tilde{\xi^+}},
$$ where $\tilde{\xi^-}$ and $\tilde{\xi^+}$ denotes some measure of
centrality of the positive and negative controls, typically the mean or
median.

The controls can themselves be passed through this function and
converted to % viability. From the variances of these normalized
controls, $\lambda$ can be set as indicated above. And the negative
controls can be added directly into the algorithm. Negative controls
represents unhindered cell growth, and can be thought of as samples from
the dose-response function $f(\mathbf{x})$ at concentration
$\mathbf{x}=(0,0)$. These can then be added directly to the
$\texttt{bayesynergy}$ function in the same way as regular observations.

# Synergy classification

Frequently, it is of interest to classify an experiment as *synergistic*
or *antagonistic*. Usually, this has been done by thresholding the
synergy measure at a certain level, declaring e.g. everything above 10
as synergistic, everything below -10 antagonistic, and anything in
between as additive (no interaction). The problem with this is that it
completely ignores the underlying measurement error, and as a
consequence the thresholding procedure can lead to misclassification.
Large synergistic effects might be classified as synergistic, but in
reality the effect cannot be discerned from the background noise. In the
same manner, genuine synergistic effects that are too small, for example
because the dose-ranges are a bit off, will also be misclassified. By
incorporating the uncertainty into the classification it can be done in
a more principled manner.

In Bayesian inference, we can compute what is know as the model
*evidence*. That is, given a probabilistic model $\mathcal{M}$, and some
data we think is generated from it, $\mathcal{D}$, the evidence is
defined as the probability of the model given the data,
$P(\mathcal{M} \vert \mathcal{D})$. We can use this quantity to compare
different models, in particular when comparing two distinct models we
can define the **Bayes Factor,** $\text{BF}_{10}$: $$
\text{BF}_{10}=\frac{P(\mathcal{D}\vert\mathcal{M}_1)}{P(\mathcal{D}\vert\mathcal{M}_0)} = \frac{P(\mathcal{M}_1 \vert \mathcal{D})}{P(\mathcal{M}_0 \vert \mathcal{D})}\frac{P(\mathcal{M}_1)}{P(\mathcal{M}_0)},
$$ where $P(\mathcal{M}_1)$ and $P(\mathcal{M}_0)$ denotes the prior
model probabilities. By defining $$
\mathcal{M}_0: f(\mathbf{x}) = p_0(\mathbf{x}) \\
\mathcal{M}_1: f(\mathbf{x}) = p_0(\mathbf{x}) + \Delta(\mathbf{x}),
$$ and computing $\text{BF}_{10}$, the Bayes factor gives information on
whether the interaction surface needs to be included in the model. A
high value indicates that $\mathcal{M}_1$ is preferred over
$\mathcal{M}_0$, and thus that there most likely is some interaction in
the experiment. One still needs to make a cutoff, but it will be less
arbitrary by connecting it directly to the uncertainty in the model, and
model evidence. The thresholding itself can be done according to e.g.
the table in @Kass1995:

```{r,echo=F,cache=T}
df = data.frame("BF10" = c("1 to 3.2","3.2 to 10","10 to 100",">100"),"Evidence against M0" = c("Not worth more than a bare mention","Substantial","Strong","Decisive"))
knitr::kable(df,col.names = c("$\\text{BF}_{10}$","Evidence against $\\mathcal{M}_0$"),align="cc",format = "simple")
```


The Bayes factor only gives information about whether or not an
interaction is present. Depending on the classification task, one still
needs to decide if the effect is synergistic or antagonistic. For this
one could e.g. use the integral of the interaction surface,
$\text{VUS}(\Delta)$, if this is negative the experiment is coded as
synergistic, if positive it is coded as antagonistic.

The calculation of the Bayes factor is implemented directly in the
`bayesynergy` function, and can be calculated simply by adding
`bayes_factor = T` to the call. Model evidence and the Bayes factor
itself is computed via the `bridgesampling` package (@Gronau2020).

# References
