---
title: "Introduction to bayesynergy"
description: >
  Learn how to get started with the basics of bayesynergy
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bayeysnergy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```


The `bayesynergy` R package implements a Bayesian semi-parametric
regression model for estimating the dose-response function of *in-vitro*
drug combination experiments. The Bayesian framework offers full
uncertainty quantification of the dose response function and any derived
summary statistics, as well as natural handling of replicates and
missing data. The Bayesian model is implemented in [RStan](https://mc-stan.org/users/interfaces/rstan),
taking advantage of the efficient 'No U-Turn Sampler' as well as
variational Bayes for quick approximations of the true posterior.

The package is further equipped with plotting functions for summarizing
a drug response experiment, parallel processing for large drug
combination screen, as well as plotting tools for summarizing and
comparing these.

## Basic usage example

To get started, simply load the package with

```{r}
library(bayesynergy)
```

We have included an example dataset from a large drug combination
screening experiment on diffuse large B-cell lymphoma. To access it,
simply run

```{r}
data("mathews_DLBCL")
y <- mathews_DLBCL[[1]][[1]]
x <- mathews_DLBCL[[1]][[2]]
head(cbind(y,x))
```

where `y` contains the measured post-treatment viability, and `x` the corresponding drug concentrations of the two drugs.

To fit the model, simply run

```{r, eval=T, echo=T, results='hide', cache=T}
fit <- bayesynergy(y,x)
```

and wait for the sampling process to complete. 

The results can then be plotted to inspect the model fit e.g. for the monotherapies

```{r, eval=T, fig.dim = c(7,5), fig.show="hold", results="hide", fig.keep = c(1)}
plot(fit)
```

And similarly for the full dose-response surface in an interactive figure

```{r 3dresponse, echo = F, cache = T, warning = FALSE, message = FALSE, fig.dim = c(7,5)}
x = fit
# Creating some stuff needed for plots
library(plotly)

  posterior = rstan::extract(x$stanfit)
  n.save = length(posterior$lp__)
  # We add lower-asymptote parameters if these are not estimated
  if (!x$model$lower_asymptotes){
    posterior$la_1 = rep(0,n.save)
    posterior$la_2 = rep(0,n.save)
  }
  unqX1 = log10(sort(unique(x$data$x[,1])))[-1] # Removing -Inf here
  unqX2 = log10(sort(unique(x$data$x[,2])))[-1] # Removing -Inf here
  dx1 = mean(diff(unqX1))
  dx2 = mean(diff(unqX2))
  nrep = ncol(as.matrix(x$data$y))
  # Need to find coordinates for the observed variables in this new coordinate system
  Xgrid = expand.grid(unqX1,unqX2)
  Xgrid = Xgrid[order(Xgrid[,"Var1"],Xgrid[,"Var2"]),]
  
  mono1 = data.frame(
    x = rep(log10(x$data$x[which((x$data$x[,2]==0) & (x$data$x[,1] != 0)),1]),nrep),
    y = as.vector(as.matrix(x$data$y)[which((x$data$x[,2]==0) & (x$data$x[,1] != 0)),])
  )
  # Remove NA here
  idx = !is.na(mono1$y)
  mono1 = mono1[idx,]
  
  mono2 = data.frame(
    x = rep(log10(x$data$x[which((x$data$x[,1]==0) & (x$data$x[,2] != 0)),2]),nrep),
    y = as.vector(as.matrix(x$data$y)[which((x$data$x[,1]==0) & (x$data$x[,2] != 0)),])
  )
  # Remove NA here
  idx = !is.na(mono2$y)
  mono2 = mono2[idx,]
  
  # Pull out indices we want
  ii = x$data$indices[which((x$data$x[,1]!=0) & (x$data$x[,2] != 0))]
  # Also define residuals here
  
  combination = data.frame(
    x1 = rep(log10(x$data$x[which((x$data$x[,1]!=0) & (x$data$x[,2]!=0)),1]),nrep),
    x2 = rep(log10(x$data$x[which((x$data$x[,1]!=0) & (x$data$x[,2]!=0)),2]),nrep),
    y = as.vector(as.matrix(x$data$y)[which((x$data$x[,1]!=0) & (x$data$x[,2] != 0)),]),
    f = 0,
    p0 = 0,
    Delta = 0,
    residuals = as.vector(as.matrix(x$data$y)[which((x$data$x[,1]!=0) & (x$data$x[,2] != 0)),]) - as.vector(x$posterior_mean$p0)[ii]
  ) 
  # Remove NA here
  idx = !is.na(combination$y)
  combination = combination[idx,]
  
  ####################################################################################
  # Monotherapies
  ####################################################################################
  grid.size = 100
  x.seq1 = seq(min(unqX1)-dx1,max(unqX1)+dx1,length.out = grid.size)
  x.seq2 = seq(min(unqX2)-dx2,max(unqX2)+dx2,length.out = grid.size)
  
  y.seq1 = matrix(NA,nrow=grid.size,ncol=n.save)
  y.seq2 = matrix(NA,nrow=grid.size,ncol=n.save)
  for (i in 1:grid.size){
    y.seq1[i,] = as.vector(posterior$la_1)+as.vector((1-posterior$la_1))/(1+10^(as.vector(posterior$slope_1)*(x.seq1[i]-as.vector(posterior$log10_ec50_1))))
    y.seq2[i,] = as.vector(posterior$la_2)+as.vector((1-posterior$la_2))/(1+10^(as.vector(posterior$slope_2)*(x.seq2[i]-as.vector(posterior$log10_ec50_2))))
  }
  df1 = data.frame(
    x = x.seq1,
    mean = apply(y.seq1,1,mean),
    median = apply(y.seq1,1,median),
    lower = apply(y.seq1,1,quantile, probs=0.025),
    upper = apply(y.seq1,1,quantile, probs=0.975)
  )
  df2 = data.frame(
    x = x.seq2,
    mean = apply(y.seq2,1,mean),
    median = apply(y.seq2,1,median),
    lower = apply(y.seq2,1,quantile, probs=0.025),
    upper = apply(y.seq2,1,quantile, probs=0.975)
  )

  # Response
    z_response = x$posterior_mean$f[-1,-1]
    fig = plot_ly(x = unqX1, y = unqX2, z = z_response)
    fig = fig %>% add_surface(cmin=0,cmax=1)
    fig = fig %>% add_trace(x = combination$x1, y = combination$x2, z = combination$y,
                            type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="black",symbol=104),name = "Observed")
    fig = fig %>% plotly::layout(scene = list(zaxis = list(range=c(min(min(0,c(mono1$y,mono2$y,combination$y))),max(max(1,c(mono1$y,mono2$y,combination$y)))),
                                                   title="% Viability",titlefont = list(size = 12)),
                                              xaxis = list(title=paste(x$data$units[1],x$data$drug_names[1]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>"),
                                              yaxis = list(title=paste(x$data$units[2],x$data$drug_names[2]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>")),
                         title = paste("Response surface:",x$data$experiment_ID,":",x$data$drug_names[1],"+",x$data$drug_names[2]))
    fig = fig %>% add_paths(x = df1$x, y = (min(unqX2)-mean(diff(unqX2))), z = df1$mean, line = list(color = "grey", dash = "dash",width=4), showlegend = F) 
    fig = fig %>% add_paths(x = (min(unqX1)-mean(diff(unqX1))), y = df2$x, z = df2$mean, line = list(color = "grey", dash = "dash",width=4), showlegend = F) 
    fig = fig %>% add_trace(x = mono1$x, y = (min(unqX2)-mean(diff(unqX2))), z = mono1$y, type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="grey",symbol=104), showlegend = F)
    fig = fig %>% add_trace(x = (min(unqX1)-mean(diff(unqX1))), y = mono2$x, z = mono2$y, type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="grey",symbol=104), showlegend = F)
    for (i in 1:length(unqX1)){
      fig = fig %>% add_trace(x = rep(unqX1[i],length(unqX1)), y = unqX2, z = z_response[,i]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    for (i in 1:length(unqX2)){
      fig = fig %>% add_trace(x = unqX1, y = rep(unqX2[i],length(unqX2)), z = z_response[i,]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    
    response_3d = fig
    
    response_3d
  
  
```

Quantitative measures of drug response and synergy, alongside summaries of the model parameters can be extracted by running
```{r}
summary(fit)
```

For a more detailed look into the functionality of the package, please see the other vignettes.



