---
title: "bayesynergy: flexible Bayesian modelling of synergistic interaction effects in in-vitro drug combination experiments"
author: "Leiv Rønneberg"
date: "05/09/2022"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{bayesynergy: flexible Bayesian modelling of synergistic interaction effects in in-vitro drug combination experiments}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 100px;
}
```


# Introduction

The `bayesynergy` R package implements a Bayesian semi-parametric
regression model for estimating the dose-response function of *in-vitro*
drug combination experiments. The Bayesian framework offers full
uncertainty quantification of the dose response function and any derived
summary statistics, as well as natural handling of replicates and
missing data. The Bayesian model is implemented in Stan (@Stan_2020),
taking advantage of the efficient 'No U-Turn Sampler' as well as
variational Bayes for quick approximations of the true posterior.

The package is further equipped with plotting functions for summarizing
a drug response experiment, parallel processing for large drug
combination screen, as well as plotting tools for summarizing and
comparing these.



# A simple example -- a single experiment

In the R package, we've attached two example datasets from a large drug
combination screening experiment on diffuse large B-cell lymphoma. We'll use these to show some simple use cases of the main functions and how to interpret the results.

Let's load in the first example and have a look at it

```{r load_example,cache=T}
library(bayesynergy)
data("mathews_DLBCL")
y = mathews_DLBCL[[1]][[1]]
x = mathews_DLBCL[[1]][[2]]
head(cbind(y,x))
```

We see that the the measured viability scores are stored in the vector
`y`, while `x` is a matrix with two columns giving the corresponding
concentrations where the viability scores were read off.

Fitting the regression model is simple enough, and can be done on
default settings simply by running the following code (where we add the
names of the drugs involved, the concentration units for plotting
purposes, and calculate the bayes factor).

```{r regular_fit, cache = T, warning = FALSE}
fit = bayesynergy(y,x, drug_names = c("ibrutinib", "ispinesib"),
                  units = c("nM","nM"),bayes_factor = T)
```

The resulting model can be summarised by running

```{r summary_fit,cache=T}
summary(fit)
```

which gives posterior summaries of the parameters of the model. 


For the complete example, please see here. (insert link)



# References
