---
title: "Single experiment example"
author: "Leiv Rønneberg"
date: "05/09/2022"
output: 
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Single experiment example}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In the R package, we've attached two example datasets from a large drug
combination screening experiment on diffuse large B-cell lymphoma. We'll use these to show some simple use cases of the main functions and how to interpret the results.

Let's load in the first example and have a look at it

```{r load_example,cache=T}
library(bayesynergy)
data("mathews_DLBCL")
y = mathews_DLBCL[[1]][[1]]
x = mathews_DLBCL[[1]][[2]]
head(cbind(y,x))
```

We see that the the measured viability scores are stored in the vector
`y`, while `x` is a matrix with two columns giving the corresponding
concentrations where the viability scores were read off.

Fitting the regression model is simple enough, and can be done on
default settings simply by running the following code (where we add the
names of the drugs involved, the concentration units for plotting
purposes, and calculate the bayes factor).

```{r regular_fit, cache = T, warning = FALSE}
fit = bayesynergy(y,x, drug_names = c("ibrutinib", "ispinesib"),
                  units = c("nM","nM"),bayes_factor = T)
```

The resulting model can be summarised by running

```{r summary_fit,cache=T}
summary(fit)
```

which gives posterior summaries of the parameters of the model. 

In addition, the model calculates summary statistics of the monotherapy
curves and the dose-response surface including drug sensitivity scores
(DSS) for the two drugs in question, as well as the volumes that capture
the notion of efficacy (`rVUS_f`), interaction (`VUS_Delta`), synergy
(`VUS_syn`) and interaction (`VUS_ant`).

As indicated, the total combined drug efficacy is around 80% (`rVUS_f`),
of which around 70 percentage points can be attributed to $p_0$
(`rVUS_p0`), leaving room for 10 percentage points worth of synergy
(`VUS_syn`). We can also note that the model is fairly certain of this
effect, with a 95% credible interval given as
(`r round(rstan::summary(fit$stanfit,probs=c(0.025,.5,0.975))$summary["VUS_syn",4],3)`,
`r round(rstan::summary(fit$stanfit,probs=c(0.025,.5,0.975))$summary["VUS_syn",6],3)`).
The certainty of this is also verified by the Bayes factor, which at
`r round(fit$bayesfactor,2)` indicates strong evidence of an interaction
effect present in the model.


# Visualization

## Monotherapy curves, 2D contour plots

We can also create plots by simply running

```{r regular_plots, cache = T, warning = FALSE, message = FALSE, fig.dim = c(7,7), fig.show="hold", results="hide", fig.keep ="all"}
plot(fit, plot3D = F)
```

which produces monotherapy curves, monotherapy summary statistics, 2D
contour plots of the dose-response function $f$, the non-interaction
assumption $p_0$ and the interaction $\Delta$. The last plot displays
the $rVUS$ scores as discussed previously, with corresponding
uncertainty.


## 3D interactive plots

The package can also generate 3D interactive plots by setting
`plot3D = T`. These are displayed as following using the plotly library
(@plotly).

### Dose-response

```{r 3dresponse, echo = F, cache = T, warning = FALSE, message = FALSE, fig.dim = c(7,7)}
x = fit
# Creating some stuff needed for plots
library(plotly)

  posterior = rstan::extract(x$stanfit)
  n.save = length(posterior$lp__)
  # We add lower-asymptote parameters if these are not estimated
  if (!x$model$lower_asymptotes){
    posterior$la_1 = rep(0,n.save)
    posterior$la_2 = rep(0,n.save)
  }
  unqX1 = log10(sort(unique(x$data$x[,1])))[-1] # Removing -Inf here
  unqX2 = log10(sort(unique(x$data$x[,2])))[-1] # Removing -Inf here
  dx1 = mean(diff(unqX1))
  dx2 = mean(diff(unqX2))
  nrep = ncol(as.matrix(x$data$y))
  # Need to find coordinates for the observed variables in this new coordinate system
  Xgrid = expand.grid(unqX1,unqX2)
  Xgrid = Xgrid[order(Xgrid["Var1"],Xgrid["Var2"]),]
  
  mono1 = data.frame(
    x = rep(log10(x$data$x[which((x$data$x[,2]==0) & (x$data$x[,1] != 0)),1]),nrep),
    y = as.vector(as.matrix(x$data$y)[which((x$data$x[,2]==0) & (x$data$x[,1] != 0)),])
  )
  # Remove NA here
  idx = !is.na(mono1$y)
  mono1 = mono1[idx,]
  
  mono2 = data.frame(
    x = rep(log10(x$data$x[which((x$data$x[,1]==0) & (x$data$x[,2] != 0)),2]),nrep),
    y = as.vector(as.matrix(x$data$y)[which((x$data$x[,1]==0) & (x$data$x[,2] != 0)),])
  )
  # Remove NA here
  idx = !is.na(mono2$y)
  mono2 = mono2[idx,]
  
  # Pull out indices we want
  ii = x$data$indices[which((x$data$x[,1]!=0) & (x$data$x[,2] != 0))]
  # Also define residuals here
  
  combination = data.frame(
    x1 = rep(log10(x$data$x[which((x$data$x[,1]!=0) & (x$data$x[,2]!=0)),1]),nrep),
    x2 = rep(log10(x$data$x[which((x$data$x[,1]!=0) & (x$data$x[,2]!=0)),2]),nrep),
    y = as.vector(as.matrix(x$data$y)[which((x$data$x[,1]!=0) & (x$data$x[,2] != 0)),]),
    f = 0,
    p0 = 0,
    Delta = 0,
    residuals = as.vector(as.matrix(x$data$y)[which((x$data$x[,1]!=0) & (x$data$x[,2] != 0)),]) - as.vector(x$posterior_mean$p0)[ii]
  ) 
  # Remove NA here
  idx = !is.na(combination$y)
  combination = combination[idx,]
  
  ####################################################################################
  # Monotherapies
  ####################################################################################
  grid.size = 100
  x.seq1 = seq(min(unqX1)-dx1,max(unqX1)+dx1,length.out = grid.size)
  x.seq2 = seq(min(unqX2)-dx2,max(unqX2)+dx2,length.out = grid.size)
  
  y.seq1 = matrix(NA,nrow=grid.size,ncol=n.save)
  y.seq2 = matrix(NA,nrow=grid.size,ncol=n.save)
  for (i in 1:grid.size){
    y.seq1[i,] = as.vector(posterior$la_1)+as.vector((1-posterior$la_1))/(1+10^(as.vector(posterior$slope_1)*(x.seq1[i]-as.vector(posterior$log10_ec50_1))))
    y.seq2[i,] = as.vector(posterior$la_2)+as.vector((1-posterior$la_2))/(1+10^(as.vector(posterior$slope_2)*(x.seq2[i]-as.vector(posterior$log10_ec50_2))))
  }
  df1 = data.frame(
    x = x.seq1,
    mean = apply(y.seq1,1,mean),
    median = apply(y.seq1,1,median),
    lower = apply(y.seq1,1,quantile, probs=0.025),
    upper = apply(y.seq1,1,quantile, probs=0.975)
  )
  df2 = data.frame(
    x = x.seq2,
    mean = apply(y.seq2,1,mean),
    median = apply(y.seq2,1,median),
    lower = apply(y.seq2,1,quantile, probs=0.025),
    upper = apply(y.seq2,1,quantile, probs=0.975)
  )

  # Response
    z_response = x$posterior_mean$f[-1,-1]
    fig = plot_ly(x = unqX1, y = unqX2, z = z_response)
    fig = fig %>% add_surface(cmin=0,cmax=1)
    fig = fig %>% add_trace(x = combination$x1, y = combination$x2, z = combination$y,
                            type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="black",symbol=104),name = "Observed")
    fig = fig %>% plotly::layout(scene = list(zaxis = list(range=c(min(min(0,c(mono1$y,mono2$y,combination$y))),max(max(1,c(mono1$y,mono2$y,combination$y)))),
                                                   title="% Viability",titlefont = list(size = 12)),
                                              xaxis = list(title=paste(x$data$units[1],x$data$drug_names[1]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>"),
                                              yaxis = list(title=paste(x$data$units[2],x$data$drug_names[2]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>")),
                         title = paste("Response surface:",x$data$experiment_ID,":",x$data$drug_names[1],"+",x$data$drug_names[2]))
    fig = fig %>% add_paths(x = df1$x, y = (min(unqX2)-mean(diff(unqX2))), z = df1$mean, line = list(color = "grey", dash = "dash",width=4), showlegend = F) 
    fig = fig %>% add_paths(x = (min(unqX1)-mean(diff(unqX1))), y = df2$x, z = df2$mean, line = list(color = "grey", dash = "dash",width=4), showlegend = F) 
    fig = fig %>% add_trace(x = mono1$x, y = (min(unqX2)-mean(diff(unqX2))), z = mono1$y, type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="grey",symbol=104), showlegend = F)
    fig = fig %>% add_trace(x = (min(unqX1)-mean(diff(unqX1))), y = mono2$x, z = mono2$y, type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="grey",symbol=104), showlegend = F)
    for (i in 1:length(unqX1)){
      fig = fig %>% add_trace(x = rep(unqX1[i],length(unqX1)), y = unqX2, z = z_response[,i]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    for (i in 1:length(unqX2)){
      fig = fig %>% add_trace(x = unqX1, y = rep(unqX2[i],length(unqX2)), z = z_response[i,]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    
    response_3d = fig
    
    response_3d
  
  
```


### Non-interaction


```{r 3dnoninteraction, echo = F, cache = T, warning = FALSE, message = FALSE, fig.dim = c(7,7)}

z_p0 = x$posterior_mean$p0[-1,-1]
    fig = plot_ly(x = unqX1, y = unqX2, z = z_p0)
    fig = fig %>% add_surface(cmin=0,cmax=1)
    fig = fig %>% add_trace(x = combination$x1, y = combination$x2, z = combination$y,
                            type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="black",symbol=104), name = "Observed")
    fig = fig %>% plotly::layout(scene = list(zaxis = list(range=c(min(min(0,c(mono1$y,mono2$y,combination$y))),max(max(1,c(mono1$y,mono2$y,combination$y)))),
                                                  title="% Viability",titlefont = list(size = 12)),
                                              xaxis = list(title=paste(x$data$units[1],x$data$drug_names[1]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>"),
                                              yaxis = list(title=paste(x$data$units[2],x$data$drug_names[2]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>")),
                         title = paste("Non-interaction surface:",x$data$experiment_ID,":",x$data$drug_names[1],"+",x$data$drug_names[2]))
    fig = fig %>% add_paths(x = df1$x, y = (min(unqX2)-mean(diff(unqX2))), z = df1$mean, line = list(color = "grey", dash = "dash",width=4), showlegend = F) 
    fig = fig %>% add_paths(x = (min(unqX1)-mean(diff(unqX1))), y = df2$x, z = df2$mean, line = list(color = "grey", dash = "dash",width=4), showlegend = F) 
    fig = fig %>% add_trace(x = mono1$x, y = (min(unqX2)-mean(diff(unqX2))), z = mono1$y, type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="grey",symbol=104), showlegend = F)
    fig = fig %>% add_trace(x = (min(unqX1)-mean(diff(unqX1))), y = mono2$x, z = mono2$y, type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="grey",symbol=104), showlegend = F)
    for (i in 1:length(unqX1)){
      fig = fig %>% add_trace(x = rep(unqX1[i],length(unqX1)), y = unqX2, z = z_p0[,i]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    for (i in 1:length(unqX2)){
      fig = fig %>% add_trace(x = unqX1, y = rep(unqX2[i],length(unqX2)), z = z_p0[i,]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    noninter_3d = fig
    noninter_3d

```

### Interaction

```{r 3dinteraction, echo = F, cache = T, warning = FALSE, message = FALSE, fig.dim = c(7,7)}

# Interaction
    z_Delta = x$posterior_mean$Delta[-1,-1]
    fig = plot_ly(type = "mesh3d")
    fig = fig %>% add_trace(x = unqX1, y = unqX2, z = z_Delta,type = "surface",
                            colorscale = list(c(0,0.5, 1), c("2166AC","EAECCC", "B2182B")),cmin=-1,cmax=1)
    fig = fig %>% add_trace(x = combination$x1, y = combination$x2, z = combination$residuals,
                            type = "scatter3d", mode = "markers",
                            marker = list(size=3,color="black",symbol=104), name = "y - p<sub>0</sub>", showlegend = T)
    fig = fig %>% plotly::layout(scene = list(zaxis= list(range=c(-1,1),
                                                  title="Interaction",titlefont = list(size = 12)),
                                              xaxis = list(title=paste(x$data$units[1],x$data$drug_names[1]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>"),
                                              yaxis = list(title=paste(x$data$units[2],x$data$drug_names[2]),titlefont = list(size = 12),tickprefix="10<sup>",tickfont=list(size=10),ticksuffix="</sup>")),
                         title = paste("Interaction surface:",x$data$experiment_ID,":",x$data$drug_names[1],"+",x$data$drug_names[2]))
    for (i in 1:length(unqX1)){
      fig = fig %>% add_trace(x = rep(unqX1[i],length(unqX1)), y = unqX2, z = z_Delta[,i]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    for (i in 1:length(unqX2)){
      fig = fig %>% add_trace(x = unqX1, y = rep(unqX2[i],length(unqX2)), z = z_Delta[i,]+0.003, type="scatter3d", mode="lines",
                              showlegend = F, line = list(color="grey", width = 1, dash = "dot"))
    }
    inter_3d = fig
    inter_3d

```

# References 


